Hierarchical and grouped time series forecasting: an application to Emergency Medical Services

### 2022-10-26 manuscript/03-experiment.qmd

# Experiment setup {#htc}

## Data

@@

@@

>>>>

# Experiment setup {#sec-experiment}

We are interested in generating forecasts to inform the planning horizon of $ph=7 to 42$ days, interested by planners in the ambulance services trust. The forecast horizon in this study is $fh = 2 \times ph$ days ahead (2*42 days planning horizon). This is because the planning is generally freezed for ph days and considering a forecast horizon of ph days might not be helpful for planning. While forecasts are generated for $2 \times ph$ days ahead, performance evaluation is only assessed based on the last $ph$ days and not the $2 \times ph$ days. The forecasts are produced for the holdout of 365 days using time series cross-validation [@hyndman2021forecasting]. 
In the following section, we discuss the dataset, describe the forecasting methods used to generate base forecasts and present the point and probabilistic accuracy measures.

## Data {#sec-data}

The dataset used in this study is from a major ambulance service trust in the United Kingdom. It contains information relating to the daily number of attended incidents from 10 October 2015 to 31 July 2019 by nature of incidents, priority, the health board managing the service and the control area (or region). Figure @fig-hierarchy depicts both the hierarchical and grouped structure of the data. Figure @fig-hierarchy-1 illustrates the nested hierarchical structure based on control area and health board and Figure @fig-hierarchy-2 shows the grouped structure by priority and the nature of incident.

```{r}
#| label: fig-hierarchy
#| out.width: "60%"
#| fig-cap: "The hierarchical and grouped structure of attended inciddents (ambulance demand)"
#| fig-subcap: 
#|   - "Hierarchical structure: Attended inciddents in all country is disaggregated into 3 control areas /regions and then into 7 different healrhboard at the bottom"
#|   - "Grouped structure:  Incidents could be grouped into priority (i.e. Red, Amber & Green) and the nature of attenedd incident (i.e. there are 35 different nature of incidents including chest pain , breahtign problems, heart attack, stroke, and son on). The symbol * refers to the crossed attributes between hierarchical and grouped levels."
#| layout: [[50],[-2], [38]]

library(ggraph)
library(igraph)
library(tidyverse)

data <- data.frame(
  level1="Total",
  level2=c("Central & West", "Central & West","Central & West","North", "South & East","South & East","South & East"),
  level3=c("HD", "SB", "PO", "BC","CV", "CT", "AB")
)
# transform it to a edge list!
edges_level1_2 = data %>% select(level1, level2) %>% 
  unique %>% rename(from=level1, to=level2)
edges_level2_3 = data %>% select(level2, level3) %>% 
  unique %>% rename(from=level2, to=level3)
edge_list=rbind(edges_level1_2, edges_level2_3)

mygraph <- graph_from_data_frame( edge_list )
ggraph(mygraph, layout = 'dendrogram', circular = FALSE) + 
  geom_edge_diagonal() +
  geom_node_point(colour="#dddddd",size=10) +
  geom_node_text(
    aes(label=c("All country", "Central & West", "North","South & East", "HD", "SB", "PO", "BC","CV", "CT", "AB") )
  )+
  theme_void()

knitr::include_graphics(here::here("img/group.png"))
```

Table @tbl-hierarchy also displays the structure of data with the total number of series at each level. At the top level, we have the total attended incidents for the country. We can split these total attended incidents by control area, by health board, by priority or by nature of incident. There are 3 control areas breakdown by 7 local health boards. Attended incident data are categorized into 3 priority classes of red, amber and green. There are also 35 different nature of incidents such as chest pain, stroke, breathing problem, etc. In total, across all levels of disaggregation, there are 1530 time series.

```{r}
#| label: tbl-hierarchy
#| tbl-cap: "Number of time series in each level for the hierarchical & grouped structure of attended incidents"
agg_level <- tibble::tribble(
  ~Level, ~`Number of series`,
  "All country", 1,
  "Control", 3,
  "Health board", 7,
  "Priority", 3,
  "Priority * Control", 9,
  "Priority * Health board", 21,
  "Nature of incident", 35,
  "Nature of incident * Control", 105,
  "Nature of incident * Health board", 245,
  "Priority * Nature of incident", 104,
  "Control * Priority * Nature of incident", 306,
  "Control * Health board * Priority * Nature of incident (Bottom level)", 691,
  "Total", 1530
)
knitr::kable(agg_level, booktabs = T, position = "left") %>% kable_classic(full_width = F)
```

Given the total number of time series, direct visual analysis is infeasible. Therefore, we first compute features of all 1530 time series and display the strength of trend and weekly seasonality strength in Figure @fig-feature. Each point represents one time series with strength of trend in  x-axis and strength of seasonality in y-axis. It is clear that there are some series showing strong trends and/or seasonality, corresponding to series at the higher levels of the hierarchy. The majority show low trend and seasonality with . These are time series belonging to the bottom of the structure, series related to the nature of incidents for each given control, health board and priority.

```{r}
#| label: fig-feature
#| cache: true
#| out.width: "70%"
#| fig.align: center
#| fig-cap: "Time series features of attended incidents across all levels (1530 series)"
library(feasts)
incident_gthf <- read_rds(here::here("data/incident_gt.rds"))
incident_gthf %>% 
  features(incidents,feat_stl) %>%
  ggplot(aes(x = trend_strength, y = seasonal_strength_week)) +
  geom_point(pch=1)+
  ggthemes::theme_few()
```

In addition to displaying the trend and seasonality features, we also highlight few time series at various levels of the aggregation. Figure @fig-dataviz reveals different information such as trend, seasonality and noise. For example, some series depict seasonality and trend, whereas some other series report low volume of attended incidents and entropy, making them more volatile and difficult to forecast. At the level on nature of incidents combined with categories of other levels, there are many series that contain zeros with low counts. As such, the data set represents a diverse set of daily time series patterns.

```{r}
#| label: fig-dataviz
#| cache: true
#| dependson: "fig-feature"
#| out.width: "70%"
#| fig-cap: "Time series of attended incidents at various levels"
#| fig-subcap: 
#|   - "all country"
#|   - "Control area"
#|   - "Health board and control area"
#|   - "Healthboard and priority"
#|   - "Nature of incident and health board"
#| layout: [[44,-2, 44], [44,-2, 44], [100]]
p_total <- incident_gthf %>%
    filter(is_aggregated(region) & is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) %>%
    autoplot()+
    labs(x="Date", y="Incidents")+
    ggthemes::theme_few()

p_total_control <- incident_gthf %>%
  filter(!is_aggregated(region) & !is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) %>%
  as_tibble() %>% select(-nature,-category) %>% 
  group_by(date,region) %>% summarise(incidents=sum(incidents), .groups="drop") %>% 
  ggplot(aes(x=date, y= incidents, colour = factor(region))) +
  geom_line()+
  labs(x="Date", y="Incidents", colour = "Control")+
  ggthemes::scale_color_colorblind()+
  ggthemes::theme_few()+
  theme(legend.position = "bottom")

p_control <- incident_gthf %>%
  filter(!is_aggregated(region) & !is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) %>%
  as_tibble() %>% select(-nature,-category) %>% 
  ggplot(aes(x=date, y= incidents, colour = factor(lhb))) +
  geom_line()+
  facet_wrap(vars(factor(region)), scales = "free_y")+
  labs(x="Date", y="Incidents", colour = "Health board")+
  ggthemes::scale_color_colorblind()+
  ggthemes::theme_few()+
  theme(legend.position = "bottom")

p_priority <- incident_gthf %>%
  filter(!is_aggregated(region) & !is_aggregated(lhb) & !is_aggregated(category) & is_aggregated(nature)) %>%
  as_tibble() %>% select(-nature,-region) %>% 
  ggplot(aes(x=date, y= incidents, colour = factor(lhb))) +
  geom_line()+
  facet_wrap(vars(factor(category)), scales = "free_y")+
  labs(x="Date", y="Incidents", colour = "Health board")+
  ggthemes::scale_color_colorblind()+
  ggthemes::theme_few()+
  theme(legend.position = "bottom")

selected_nature <- c("CHESTPAIN", "STROKECVA", "BREATHING", "ABDOMINAL")

p_nature_healthboard <- incident_gthf %>%
  filter(!is_aggregated(region) & !is_aggregated(lhb) & !is_aggregated(category) & !is_aggregated(nature)) %>% 
as_tibble() %>% mutate(nature=as.character(nature)) %>% 
  filter(nature %in% selected_nature) %>% group_by(date,nature,lhb) %>% summarise(incidents=sum(incidents), .groups="drop") %>% 
  ggplot(aes(x=date, y= incidents, colour = factor(lhb))) +
  geom_line()+
  facet_wrap(vars(factor(nature)), scales = "free_y")+
  labs(x="Date", y="Incidents", colour = "Health board")+
  ggthemes::scale_color_colorblind()+
  ggthemes::theme_few()+
  theme(legend.position = "none")

p_total
p_total_control
p_control
p_priority
p_nature_healthboard
```

@@

Given the presence of various significant patterns in the past attended incidents, we consider three different forecasting models to generate the base forecasts. Once the base forecasts are produced, hierarchical and grouped time series methods are used to reconcile them across the all levels. We briefly discussed forecasting models in the following sections, and the hierarchical forecasting methods are discussed in @sec-htc.

**Exponential Smoothing State Space model (ETS):** ETS models [@hyndman2021forecasting] can combine trend, seasonality and error components in a time series through various forms such as additive, multiplicative or mixed. The trend component can be none ("N"), Additive ("A"), damped ("Ad") or multiplicative ("M"). The seasonality can be none ("N"), Additive ("A"), or multiplicative ("M"). The error term can also be additive ("A") or multiplicative ("M"). To forecast the attended incidents at each level, we use the `ETS()` function in the `fable` package [@fable2022] in R. To identify the best model for a given time series, the ETS function uses the corrected Akaike’s Information Criterion (AICc).

**Auto-Regressive Integrated Moving Average (ARIMA):** ARIMA [@hyndman2021forecasting] is another effective family of forecasting models that capture autocorrelation in time series. They can model various range of time series considering the stationarity of series, the Auto-Regressive (AR) and moving average (MA) orders. In this study, an automatic ARIMA algorithm implemented using `ARIMA()` function of fable package in R is used to forecast the attended incidents. To select the best model for a given time series, the algorithm uses unit root test, Maximum Likelihood Estimation (MLE) and the AICc.

**Poisson Regression: ** Despite the popularity and the relevant of automatic `ETS` and `ARIMA` in this study, but they produce non-integer attended incidents and forecasts might also be negative. However, the number of attended incidents is an integer and non-negative. When using `ETS` and `ARIMA`, a time series transformation approach such as `sqrt()` could be used to generate strictly positive forecasts, however forecasts are still not integer. Another alternative is to use  forecasting models that produce integer, non-negative forecasts. One of the models that is frequently used in practice is the Poisson regression [see, for example, @mccarthy2008challenge], which can be summarized as
$$
  \begin{aligned}
    & {y}_t \sim \mathrm{Poisson} \left( \lambda_t \right) \\
    & \log \lambda_t = \mathbf{x}_t' \boldsymbol{\beta}
  \end{aligned} \quad
$$ {#eq-poisson-regression}

The logarithm in @eq-poisson-regression is needed to ensure that the parameter of Poisson distribution is always positive. This model can be estimated via maximization of the likelihood function based on Poisson mass function. For taking into account serial dependence we include a regression on the previous observation. We try to capture the short range serial dependence by a first order autoregressive term and the weekly seasonality by a 7th order autoregressive term. We use the `tscount()` function in the fable package in R, which is a wrapper function written based on `tscount` package. We fit a model to each time series using the function tsglm, with the identity link function, defined by the argument link.

@@

Forecasting performance is evaluated using both point and probabilistic error measures.

The point forecast accuracy is measured via Root Mean Squared Scaled Error (RMSSE):

$$
\text{RMSSE} = \sqrt{\text{mean}(q_{j}^2)},
$$ {#eq-RMSSE}

where,

$$
q^2_{j} = \frac{\displaystyle e^2_{j}}
    {\displaystyle\frac{1}{T-m}\sum_{t=m+1}^T (y_{t}-y_{t-m})^2},
$$

$e_{j}$ is the point forecast error $j$ and $m = 1$ for non-seasonal series and $m = 7$ for daily seasonal series, $y_t$ is the observation for period $t$ and $T$ is the sample size (observations used for training the forecasting model). Smaller RMSSE values suggest more accurate forecasts. Note that the measure is scale-independent, thus allowing us to average the results across series.

To measure the forecast distribution performance, we calculate  the Continuous Rank Probability Score [@gneiting2014probabilistic]. It rewards sharpness and penalizes miscalibration, so it measures overall performance.

$$
CRPS=\frac{1}{h} \sum_{j=1}^{h} \int_{-\infty}^{\infty} \left(F^f_j(x) - F^0_j(x)\right)^2dx
$$ {#eq-CRPS}

where $F^f_j(x)$ is the forecasted Cumulative Density Function (CDF) of period $j$ and 𝐹$F^0_j(x)$ is the true CDF of period $j$.

### 2022-10-26 manuscript/04-htsgts.qmd

# Hierachcail and grouped time series forecasting techniques {#htc}

Moreover, both models either underestimated or overestimated the number of calls. Authors suggested that the inability to capture a positive trend resulted in underestimation, and the opposite was due to a negative trend. Therefore, authors recommended that further research should add certain temporal variables able to capture the trend.

## Notation

## Base forecast

## Bottom-up

## Optimal reconciliation

>>>>

# Hierachcail and grouped time series forecasting techniques {#sec-htc}

There are many applications in the healthcare and in particular in EMS where a collection of time series is available. These series are generally hierarchically organized based on multiple levels such as area/region, health board and/or are aggregated at different levels in groups based on nature of demand, priority of demand or some other attributes. While series could be strictly hierarchical or only grouped bases on some attributes, in many situation a more complex structures arise when attributes of interest are both nested and crossed, having hierarchical and grouped structure. This is also the case as discussed in @sec-data

## Independent (base forecast)

A common practice in healthcare (and EMS) to predict hierarchical and grouped series relies on producing independent forecast, also refereed to as base forecast, typically by different teams as the need for such forecasts arise. This is also known as base forecast. We observe $n$ time series at time $t$, across the entire hierarchical and grouped structure, written as $y_t$. The base forecasts of $y_{T+h}$ given data $y_1,\dots,y_T$ are denoted by $\hat{y}_h$ for h steps-ahead for all $n$ series ($n=1530$ in this study). Forecasts generated in this way are nor coherent.

## Reconciliation methos

Traditionally, alternative approaches to produce coherent forecasts for hierarchical and grouped time series involves using bottom-up and top-down methods by generating forecasts at a single level and then aggregate or disaggregate. Top-down requires having a unique structure to disaggregated forecasts generated at top level by proportions. However, given that we have multiple grouped attributes combined with the hierarchical structure, there is no unique way to disaggregate top forecast. Hence the top-down can not be used in this case, so either we can do some kind of reconciliation or must define our own top-down method for each hierarchy. The recommended approach is to use reconciliation.
In the following sections, we first discuss some notations and then present bottom-up and forecast reconciliation approach used in this study to generate coherent forecasts.

### Notations

Let $b_t$ be a vector of $n_b$ "bottom-level" time series at time $t$, and let $a_t$ be a corresponding vector of $n_a = n-n_b$ aggregated time series, where

$$
 a_t = Ab_t
$$

and $A$ is the $n_a\times n_b$ "aggregation" matrix specifying how the bottom-level series $b_t$ are to be aggregated to form $a_t$. The full vector of time series is given by
$$
 y_t = \begin{bmatrix}a_t \\b_t\end{bmatrix}.
$$
This leads to the $n\times n_b$ "summing" or "structural" matrix given by
$$
  S = \begin{bmatrix}A \\ I_{n_b}\end{bmatrix}
$$
such that $y_t = Sb_t$.

### Bottom-up (BU) and linear reconciliation methods

Bottom-Up is a simple approach to generate coherent forecasts. It first involves creating the base forecasts for the bottom level series. These forecasts are then aggregated to the upper levels which results in generating coherent forecasts. BU approach can capture the dynamics of the series at the bottom level, but they may be noisy and difficult to forecast as well. BU approach is limited on using only forecasts at the bottom level and does not utilize all the information available across the hierarchical and grouped structure. Forecast reconciliation approaches fill this gap and combine and reconcile all the base forecasts in order to produce coherent forecasts.

Given the summing matrix and base forecasts, bottom-up and linear reconciliation methods can be written as $\tilde{y}_h = SG\hat{y}_h$ for different matrices $G$.

Optimal reconciled forecasts are obtained with $G=(S'W^{-1}S)^{-1}W^{-1}$, or $\tilde{y}_h = M\hat{y}_h,$ where the $n\times n$ "mapping" matrix is given by $M = S(S'W^{-1}S)^{-1}W^{-1},$ where $\hat{y}_h$ are the $h$-step forecasts of $y_{T+h}$ given data to time $T$, and $W$ is an $n \times n$ positive definite matrix. Different choices for $W$ lead to different solutions such as Ordinary Least Square (OLS) , Weighted Least Square (WLS) and Minimum Trace (MinT) [@WicEtAl2019]. We use the implementation of these methods in the fable package in R in the experiment.
